```{r knitsetup, echo=FALSE, results='hide', warning=FALSE, message=FALSE, cache=TRUE}
opts_knit$set(base.dir='./', fig.path='', out.format='md')
opts_chunk$set(prompt=TRUE, comment='', results='markup')
# See yihui.name/knitr/options for more Knitr options.
##### Put other setup R code here


# end setup chunk
```
# Day 2 Session 1
---

### Session 4 - Writing a Report in RMarkdown

Using the codes we learned so far, let's compile these into a cohesive report.

1. Create an RMarkdown file - go to the menu "File"/"New File"/"RMarkdown". For now let's create a HTML formatted file (`*.HTML`). We'll run through this again at the end and also make a MS Word document (`*.DOC`).

2. This will create a template that we'll experiment with.

3. We'll be adding code chunks to perform certain actions and the output will be captured in our report.

4. Read in the dataset

```
data.csv <- read.csv(file="Dataset_01_comma.csv")
data.csv
```

```{r echo=FALSE}
data.csv <- read.csv(file="C:/MyGithub/CDCRworkshop/datasets/Dataset_01_comma.csv")
kable(data.csv)

```

5. Correct the typo for height, compute the BMIs and make the histogram plot

```
data.csv[18,"Height"] <- 5.6
data.csv$bmiPRE <- (data.csv$WeightPRE*703)/((data.csv$Height*12)**2)
data.csv$bmiPOST <- (data.csv$WeightPOST*703)/((data.csv$Height*12)**2)
hist(data.csv$bmiPRE, freq=FALSE)
lines(density(data.csv$bmiPRE))
```

```{r plot1d2s4, echo=FALSE}
data.csv[18,"Height"] <- 5.6
data.csv$bmiPRE <- (data.csv$WeightPRE*703)/((data.csv$Height*12)**2)
data.csv$bmiPOST <- (data.csv$WeightPOST*703)/((data.csv$Height*12)**2)
hist(data.csv$bmiPRE, freq=FALSE)
lines(density(data.csv$bmiPRE))

```


6. Let's add our custom summary stats. We'll also save the output and use `kable()` from the knitr package to make a prettier table in markdown format.

```
mystats <- function(x, na.omit=FALSE){
  if(na.omit)
    x <- x[!is.na(x)]
  m <- mean(x)
  n <- length(x)
  s <- sd(x)
  md <- mad(x)
  q5 <- quantile(x, .05)
  q25 <- quantile(x, .25)
  q75 <- quantile(x, .75)
  q95 <- quantile(x, .95)
  return(c(n=n, mean=m, stdev=s, MAD=md, 
           q5, q25, 
           q75, q95))
}

out <- sapply(data.csv[,c(9,10)], mystats)
kable(out)
```

```{r echo=FALSE}
mystats <- function(x, na.omit=FALSE){
  if(na.omit)
    x <- x[!is.na(x)]
  m <- mean(x)
  n <- length(x)
  s <- sd(x)
  md <- mad(x)
  q5 <- quantile(x, .05)
  q25 <- quantile(x, .25)
  q75 <- quantile(x, .75)
  q95 <- quantile(x, .95)
  return(c(n=n, mean=m, stdev=s, MAD=md, 
           q5, q25, 
           q75, q95))
}

out <- sapply(data.csv[,c(9,10)], mystats)
kable(out)
```


6. Let's also add the linear fits to a plot of the POST vs PRE BMI values

```
plot(data.csv$bmiPRE, data.csv$bmiPOST, "p")
abline(lm(data.csv$bmiPOST ~ data.csv$bmiPRE), col="red")
lines(lowess(data.csv$bmiPRE, data.csv$bmiPOST), col="blue")
```

```{r plot2d2s4, echo=FALSE}
plot(data.csv$bmiPRE, data.csv$bmiPOST, "p")
abline(lm(data.csv$bmiPOST ~ data.csv$bmiPRE), col="red")
lines(lowess(data.csv$bmiPRE, data.csv$bmiPOST), col="blue")
```


7. Linear fit output

```
fit1 <- lm(bmiPOST ~ bmiPRE, data=data.csv)
fit1
summary(fit1)
coef(fit1)
anova(fit1)
```

```{r echo=FALSE}
fit1 <- lm(bmiPOST ~ bmiPRE, data=data.csv)
fit1
summary(fit1)
coef(fit1)
anova(fit1)
```







